The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-30 article_final/main.tex
Some small changes in figure captions

After that in \sectionref{sec:experiments} compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT and GPT-2 Medium on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories dataset \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a \emph{single} matrix (the connection to the general case is presented in \sectionref{sec:lmo_for_nn}):
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot} = \<\cdot,\cdot>$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mM}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mM,\mD>\,.
\end{equation}

@@

To derive the update formula for arbitrary $k$, we use the expression for the norm dual to the Ky Fan $k$-norm (see, for instance \citep{bhatia2013matrix}, p. 96):
\[
    \normkfk{\cdot}^\dagger = \max\left\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\right\}\,.
\]
According to \equationref{eq:our_update}, we need to find $\mD$ such that $\normkfk{\mD} = 1$ and
\[
\<\mM^t, \mD> = \max\left\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i,\ \sigma_1\right\}\,.
\]
The Neon update $\mD = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\mD = \frac{1}{k}\mU\mV^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which of those two expressions achieves smaller residual on the linear function $L(\mX) := F(\mX^t) + \<\mM^t, \mX - \mX^t>$. The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

Similarly, combining Fanion-$k$ with SignSGD yields \emph{S-Fanion-$k$} with update
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where \texttt{sign\_lr\_coeff} $\eta$ is a scaling coefficient specific to SignSGD.

By \corollaryref{corollary:lin_comb_alg}, this defines an LMO-algorithm with norm $\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger$, where
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}} = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.sec:lmo_for_nn
\end{equation}

@@

Next, we tune F-Muon specifically with $\alpha = 0.5$, obtaining optimal hyperparameters \texttt{lr=0.4(1 - step/\allowbreak{}total\_steps)}, \texttt{momentum=0.6}, \texttt{nesterov = True}. Tuned F-Muon achieves $94.02 \pm 0.13\%$ validation accuracy after 8 epochs (averaged over 200 runs), matching Muon's accuracy and variance. We again measure the validation accuracy as a function of $\alpha$ (\figureref{fig:fmuon_alphas}) and observe that even at $\alpha=0.1$, the accuracy substantially exceeds that of vanilla NSGD.

For S-Muon with $\alpha=0.5$, we tune all hyperparameters to find the optimal configuration \texttt{lr=0.42(1 - step/\allowbreak{}total\_steps)}, \texttt{momentum=0.63}, \texttt{nesterov = True}, \texttt{sign\_lr\_coeff = 0.003}, achieving $94.03 \pm 0.13\%$ validation accuracy, which slightly exceeds vanilla Muon's performance.

@@

We fine-tuned the NanoGPT framework by Karpathy \citep{Karpathy2022} using the standard \emph{NanoGPT-medium} configuration, which corresponds to GPT-2 Medium. This model consists of 24 transformer layers, a hidden dimension of 1024, 16 attention heads, and approximately 345 million parameters. We selected the \emph{Tiny Stories} corpus \citep{eldan2023tinystoriessmalllanguagemodels} as the training dataset for its high entropy and structural diversity, which amplify the differences in optimization dynamics. All training runs were initialized with the same random seed, weight initialization, and learning rate schedule to ensure that performance differences arose solely from the choice of optimizer.

@@

A common limitation of these analyses is their focus on convergence measured by the gradient norm. As we showed in our CIFAR experiments \cref{sec:cifar_underhood}, the gradient norm may decrease by only a factor of ten when the accuracy reaches 100\%.

@@

\appendix
\section{LMO for Neural Networks}\label{sec:lmo_for_nn}

@@

@@

Based on the \lemmaref{lemma:dual_to_conv_comb}, we immediately find $\normfstar{\cdot}^\dagger$, which is related to F-Muon update. Indeed, after setting $\beta=1-\alpha$ and remembering that for smooth and bounded cases we can use $\min$ instead of $\inf$, we get
\begin{equation}\label{eq:normfstardual}
  \normfstar{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \norms{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

>>>>

After that in \sectionref{sec:experiments} compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT on FineWeb dataset \citep{modded_nanogpt_2024}
  \item Pre-training GPT-2 medium 
	\item Fine-tuning NanoGPT on Tiny Stories dataset \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a \emph{single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot} = \<\cdot,\cdot>$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mM}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mM,\mD>\,.
\end{equation}

@@

To derive the update formula for arbitrary $k$, we use the expression for the norm dual to the Ky Fan $k$-norm (see, for instance \citep{bhatia2013matrix}, p. 96):
\[
    \normkfk{\cdot}^\dagger = \max\left\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\right\}\,.
\]
According to \equationref{eq:our_update}, we need to find $\mD$ such that $\normkfk{\mD} = 1$ and
\[
\<\mM^t, \mD> = \max\left\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i,\ \sigma_1\right\}\,.
\]
The Neon update $\mD = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\mD = \frac{1}{k}\sum_{i=1}^{\min(m,n)}\vu_i \vv_i^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which of those two expressions achieves smaller residual on the linear function $L(\mX) := F(\mX^t) + \<\mM^t, \mX - \mX^t>$. The Ky Fan norms thus fail to close the rank gap: the resulting update is either rank-one or full-rank.

@@

Similarly, combining Fanion-$k$ with SignSGD yields \emph{S-Fanion-$k$} with update
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where $\eta$ is a scaling coefficient specific to SignSGD.

By \corollaryref{corollary:lin_comb_alg}, this defines an LMO-algorithm with norm $\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger$, where
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}} = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.
\end{equation}

@@

Next, we tune F-Muon specifically with $\alpha = 0.5$, obtaining optimal hyperparameters \texttt{lr=0.4(1 - step/total\_steps), momentum=0.6, nesterov = True}. Tuned F-Muon achieves $94.02 \pm 0.13\%$ validation accuracy after 8 epochs (averaged over 200 runs), matching Muon's accuracy and variance. We again measure the validation accuracy as a function of $\alpha$ (\figureref{fig:fmuon_alphas}) and observe that even at $\alpha=0.1$, the accuracy substantially exceeds that of vanilla NSGD.

For S-Muon with $\alpha=0.5$, we tune all hyperparameters to find the optimal configuration \texttt{lr=0.42(1 - step/total\_steps), momentum=0.63, nesterov = True, sign\_lr\_coeff = 0.003}, achieving $\mathbf{94.03} \pm 0.13\%$ validation accuracy, which slightly exceeds vanilla Muon's performance.

@@

We fine-tuned the NanoGPT framework by Karpathy \citep{Karpathy2022} using the standard \textit{NanoGPT-medium} configuration, which corresponds to GPT-2 Medium. This model consists of 24 transformer layers, a hidden dimension of 1024, 16 attention heads, and approximately 345 million parameters. We selected the \textit{Tiny Stories} corpus \citep{eldan2023tinystoriessmalllanguagemodels} as the training dataset for its high entropy and structural diversity, which amplify the differences in optimization dynamics. All training runs were initialized with the same random seed, weight initialization, and learning rate schedule to ensure that performance differences arose solely from the choice of optimizer.

@@

A common limitation of these analyses is their focus on convergence measured by the gradient norm. As we showed in our CIFAR experiments, the gradient norm may decrease by only a factor of ten when the accuracy reaches 100\%.

@@

\appendix
\section{LMO for Neural Networks}

@@

The update rule used in Muon optimizer is uSCG \citep{pethick2025training}. In the most general case, which involves momentum, it can be written as
\begin{equation}\label{eq:usgd_update}
  \begin{aligned}
    \mM^{t}     & = \alpha_{t} g(\mX^t, \xi^t) + (1 - \alpha_t)\mM^{t - 1}\,, \\
    \mX^{t + 1} & = \mX^t + \gamma_t\mathrm{LMO}(\mM^{t})\,.
  \end{aligned}
\end{equation}

Based on the lemma, we immediately find $\normfstar{\cdot}^\dagger$, which is related to F-Muon update. Indeed, after setting $\beta=1-\alpha$ and remembering that for smooth and bounded cases we can use $\min$ instead of $\inf$, we get
\begin{equation}\label{eq:normfstardual}
  \normfstar{\mY}^\dagger = \min_{\mZ} \min_t\{t, s.t. \norms{\mZ}\leq \alpha t, \normf{\mY-\mZ}\leq (1-\alpha) t\}
\end{equation}

@@
